Deep Reinforcement Learning for Conservation Decisions

### 2021-04-15 manuscript/appendix.Rmd
:pen:

@@

## Python dependencies loaded via R
sb3         <- import ("stable_baselines3")
gym         <- import ("gym")
gym_fishing <- import("gym_fishing")
gym_conservation <- import("gym_conservation")

## reproducible settings
np <- import("numpy")
np$random$seed(42L)
# Set to "" to force CPU-evaluation if needing perfect reproducibility
Sys.setenv("CUDA_VISIBLE_DEVICES"="")

@@

@@

We train the A2C algorithm using default hyper-parameter settings, based on the original publication [@A2C; @SB3].  While default hyperparameters provide a useful starting place (particularly when the environment has been suitably normalized, a best-practice we discuss above), better performance can almost always be achived by _tuning_ the hyperparameter selection.

```{r eval=!file.exists("cache/a2c.zip")}
# train an agent (model) on one of the environments:
a2c <- sb3$A2C('MlpPolicy', env, verbose=0L, seed = 42L) # Must use L for integers!
a2c$learn(total_timesteps=300000L)
a2c$save("cache/a2c")
```

```{r}
# Simulate management under the trained agent
a2c <- sb3$A2C$load("cache/a2c")
a2c_sims <- env$simulate(a2c, reps = 500L)
a2c_policy <- env$policyfn(a2c, reps = 50L)
```

Hyperparameters are selected through the process of model tuning, see.

```{r}
# train an agent (model) on one of the environments:
hyper = list('gamma'= 0.995, 'lr'= 0.0001355522450968401, 'batch_size'= 128L, 
         'buffer_size'= 10000L, 'episodic'= FALSE, 'train_freq'= 128L, 
         'noise_type'= 'normal', 'noise_std'= 0.6656948079225263, 
         'net_arch'= 'big')
policy_kwargs = list(net_arch=c(400L, 300L)) # "big"

#non-episodic:
hyper['gradient_steps'] = hyper['train_freq']
hyper['n_episodes_rollout'] = -1 # deprecated in latest SB3
n_actions = env$action_space$shape[0]
hyper$action_noise = sb3$common$noise$NormalActionNoise(
        mean=np$zeros(n_actions),
        sigma= hyper[['noise_std']] * np$ones(n_actions)
        )

td3 = sb3$TD3( 'MlpPolicy', 
               env,  
               verbose=0L, 
               seed = 42L,
               policy_kwargs=policy_kwargs,
               learning_rate = hyper[['lr']],
               gamma = hyper[['gamma']],
               batch_size = hyper[['batch_size']],            
               buffer_size = hyper[['buffer_size']],
               action_noise = hyper[['action_noise']],
               train_freq = hyper[['train_freq']],
               gradient_steps = hyper[['train_freq']],
               n_episodes_rollout = hyper[['n_episodes_rollout']]
              )
```

```{r eval=!file.exists("cache/td3.zip")}
td3$learn(total_timesteps=300000L)
td3$save("cache/td3")
```

@@

>>>>

# Deep Reinforcement Learning Frameworks

At the time of writing, several major frameworks exist which provide convenient access to baseline algorithms commonly used in deep reinforcement learning.

# Deep Reinforcement Learning in R

All though all the necessary tooling for RL is implemented in python, the R language is more familiar to most ecologists.
Fortunately, modern bindings such as the `reticulate` package [@reticulate] make it straight forward to use these tools without ever leaving the R interface.
In this appendix, we detail this "pure R" approach, as well as a "pure python" approach.

In the R based-approach, R functions take responsibility from the user for translating commands into python code before it is executed, an approach commonly referred to as meta-programming.
This still requires a local python installation, which can be installed directly from R using the command `install_miniconda()` from the `reticulate` package in R.
Alternately, users may prefer running the analysis inside a docker container.  
The Rocker Project [@rocker] provides pre-built docker containers which include the necessary R and python environments, as well as CUDA libraries required to take advantage of GPU-based acceleration on suitable architectures.
This can be particularly useful for users running ML algorithms on remote servers, as configuring R, python, and CUDA environments appropriately can be challenging.

Clone the repository <https://github.com/boettiger-lab/rl-intro>, e.g. using the New Project->From Version Control->Git menu in RStudio.
From the project directory, we can then install all the necessary dependencies using `renv`, which helps ensure a reproducible environment of fixed versions of R packages and python modules.

```{r}
#install.packages("renv")
renv::restore()
```

Once the packages have installed, we are ready to load the necessary libraries.  Note that the `import` function from `reticulate` package acts much like the `library` command in R, though it does not attach the package function to the global namespace.  To make it more convenient to access those functions, we can assign a shorthand name.

@@

## Python dependencies loaded via R
sb3         <- import ("stable_baselines3")
gym         <- import ("gym")
gym_fishing <- import("gym_fishing")
gym_conservation <- import("gym_conservation")
```

Numerical reproducibility can be challenging in machine learning problems, particularly when using GPU-based acceleration. In addition to setting a random seed in our python environment, we can optionally disable GPU use to improve reproducibility by setting the `CUDA_VISIBLE_DEVICES` to a blank value.

```{r}
## reproducible settings
np <- import("numpy")
seed <- 24L # integer
np$random$seed(seed)
# Set to "" to force CPU-evaluation if needing perfect reproducibility
Sys.setenv("CUDA_VISIBLE_DEVICES"="")

@@

We begin by select an environment from `gym_fishing` by name, passing the optional parameter `sigma` to initialize the environment with multiplicative Gaussian environmental noise of sd of `0.1`.

@@

At the heart of each algorithm is a neural network. The type, size and depth of network, choice of activation functions and so forth, are considered "hyper-parameters" of algorithm: 
we must first make a selection for each of these hyper-parameters to define our (untrained) agent.  
Other hyper-parameters determine other aspects of the training regime: such as how the agent chooses to explore possible actions, the learning rate, or how future rewards are discounted (gamma).
Each algorithm may have different hyper-parameters. `stable-baselines3` provides default values for all hyperparameters based on the original papers that introduced the corresponding algorithms,
for instance, the A2C algorithm originally described in @A2C. 
Here, we train an agent using the A2C algorithm using default hyperparameter settings.

```{r}
a2c <- sb3$A2C('MlpPolicy', env, verbose=0L, seed = seed) # L indicates a (Long) integer, not floating point
```

Training is the main computationally intensive process, which can take anywhere from a few minutes to many days, depending on the complexity of the environment and the number of training iterations budgeted.
Therefore, we save the trained agent to disk, and only execute the training method (`learn`) if no previously saved agent is found:

```{r}
if(!file.exists("cache/a2c.zip")) {

  # Train the agent for a fixed number of timesteps
  a2c$learn(total_timesteps=300000L)

  # Save our trained agent for future use
  a2c$save("cache/a2c")
}
```

Note that while default hyperparameters provide a useful starting place (particularly when the environment has been suitably normalized, a best-practice we discuss above), better performance can almost always be achieved by _tuning_ the hyperparameter selection. This is discussed further below.

We can load a trained agent and use `simulate` and `policyfn` methods provided by the `gym_fishing` environment to simulate the process for a fixed number of replicates.

```{r}
# Simulate management under the trained agent
a2c <- sb3$A2C$load("cache/a2c")
```

## represent the initial state size in the 'rescaled' state space.
state <- env$get_state( 0.75 )

for(i in 1:100){

  out <- a2c$predict(state)
  action <- out[[1]]
  result <- env$step(action)
  state <- result[[1]]

}

```{r}
a2c_sims <- env$simulate(a2c, reps = 500L)
```

```{r}
a2c_policy <- env$policyfn(a2c, reps = 50L)
```

## TD3

We train a second agent based on the TD3 algorithm.  
This time, we will declare our choices for the hyper-parameters explicitly.
Note that some of these hyperparameters, such as the `learning_rate` and discount rate, `gamma`, are also hyper-parameters of the A2C algorithm, while others, like the `action_noise` describing how new actions are proposed, are not used in A2C.
Good choices for hyperparameters can be found through a process known as "hyper-parameter tuning", which simply uses standard function optimization approaches to vary the choice of hyper-parameters used in training to determine which maximizes the agent's performance. 
For simplicity, we show only training with specified hyper-parameters here, example scripts for tuning the algorithm to discover these values can be found in <https://github.com/boettiger-lab/rl-toolkit>, and compare to the tuning utilities found in `stable-baselines3-zoo` [@zoo].

```{r}
# train an agent (model) on one of the environments:

policy_kwargs = list(net_arch=c(400L, 300L)) # "big"
# non-episodic action noise:
noise_std = 0.6656948079225263
n_actions = env$action_space$shape[0]
action_noise = sb3$common$noise$NormalActionNoise(
        mean=np$zeros(n_actions),
        sigma= noise_std * np$ones(n_actions)
        )

td3 = sb3$TD3( 'MlpPolicy', 
               env,  
               verbose=0L, 
               seed = seed,
               policy_kwargs = policy_kwargs,
               learning_rate = 0.0001355522450968401,
               gamma= 0.995, 
               batch_size = 128L,
               buffer_size = 10000L,
               train_freq = 128L, 
               n_episodes_rollout = -1, 
               gradient_steps = 128L,
               action_noise = action_noise
              )
```

Once again, after our agent is defined we are ready to train it.

```{r }
if(!file.exists("cache/td3.zip")){

  td3$learn(total_timesteps=300000L)
  td3$save("cache/td3")

}
```

@@

noise_std = 0.4805935357322933
action_noise = OrnsteinUhlenbeckActionNoise(mean = np.zeros(1),
                                            sigma = noise_std * np.ones(1))

model = TD3('MlpPolicy', 
            env, 
            verbose = 0, 
            seed = seed,
            "gamma": 0.995,
            "learning_rate":  8.315382409902049e-05,
            "batch_size": 512,
            "buffer_size": 10000,
            "train_freq": 1000,
            "gradient_steps": 1000,
            "n_episodes_rollout": -1,
            "action_noise": action_noise,
            "policy_kwargs": list("net_arch": c(64,64)))

```{r}
model$learn(total_timesteps = 1500000)

### 2021-04-15 manuscript/manuscript.Rmd
:pen:

@@

@@

We provide three examples that illustrate the application and potential of deep RL to ecological and conservation problems, highlighting both the potential and challenges there-in.
The first example focuses on the well-studied problem of setting harvest quotas in fisheries management [@Schaefer1954; @Clark1974; @Reed1979; @Worm2006; @Costello2018].
Given an appropriately estimated model, the optimal harvest policy can be determined precisely (e.g. by stochastic dynamic programming, SDP, see @Marescot2013), and under most circumstances takes the form of a "constant-escapement" policy [see mathematical proof in @Reed1979].
This provides a natural first benchmark for deep RL approaches, since we can compare the RL solution to the mathematical optimum directly.
The problem is also of interest because the RL approach can be extended to consider ecological dynamics that either surpass the complexity possible with SDP methods, or violate assumptions of such classical approaches).
Moreover, fisheries management is both an important global challenge in it's own right [@Worm2006; @Costello2018] as well as a common test case to understand ecological dynamics and conservation management more broadly [@].

For illustrative purposes, we consider the simplest version of the dynamic optimal harvest problem as outlined by @Clark1974 or @Reed1979 (covering the stochastic case).
In short, the manager seeks to optimize the net present value (discounted cumulative catch) of a fishery, observing the stock size each year and setting an appropriate harvest quota in response.
In the classical approach, the best model of the fish population dynamics must first be estimated from data, potentially with posterior distributions over parameter estimates reflecting any uncertainty.
From this model, the optimal harvest policy -- that is, the a function which returns the optimal quota for each possible observed stock size -- can be determined either by analytic [@Reed1979] or numerical [@Marescot2013] methods, depending on the complexity of the model.
In a RL approach, the model is instead replaced by a computer simulation, usually referred to as an "environment."  The simplest simulation is merely an implementation of the estimated model itself.
However, unlike an analytic or SDP solution that is derived directly from the estimated model, the RL approach is in principle agnostic to the details of the simulation.

@@

## Beyond classic methods: Non-stationary models

>>>>

<!--

Conceptual figure:

Algorithm 
  hyper-parameters specifying neural network, learning rate, etc.  
  We refer to a particular selection of algorithm and hyperparameters as an "agent", which is then "trained" to solve one or more environments
  internal 'parameters', estimated by "training" the agent.

-->

@@

We provide three examples that illustrate the application and potential of deep RL to ecological and conservation problems, highlighting both the potential and challenges there-in.

@@

The first example focuses on the important but well-studied problem of setting harvest quotas in fisheries management.
Determining fishing quotas is both a critical ecological issue [@Worm2005; @Worm2009; @Costello2018], and also a textbook example that has long informed the management of renewable resources within fisheries and beyond [@Clark1990].
Global fisheries represent an essential source of nutrition and natural diversity worldwide with [@Worm2006; @Costello2018]
Given a population growth model that predicts the total biomass of a fish stock in the following year as a function of the current biomass, it straight forward to determine what biomass corresponds to the maximum growth rate of the stock, or $B_{\textrm{MSY}}$, the biomass at Maximum Sustainable Yield (MSY) [@Schaefer1954].
When the population growth rate is stochastic, the problem is slightly harder to solve, as the harvest quota must constantly adjust to the ups and downs of stochastic growth, but it is still possible to show the optimal strategy merely seeks to maintain the stock at $B_{\textrm{MSY}}$, adjusted for any discounting of future yields [@Reed1979].

This provides a natural first benchmark for deep RL approaches, since we can compare the RL solution to the mathematical optimum directly.
The problem is also of interest because the RL approach can be extended to consider ecological dynamics that either surpass the complexity possible with SDP methods, or violate assumptions of such classical approaches).
Moreover, fisheries management is both an important global challenge in it's own right [@Worm2006; @Costello2018] as well as a common test case to understand ecological dynamics and conservation management more broadly [@].

For illustrative purposes, we consider the simplest version of the dynamic optimal harvest problem as outlined by @Clark1974 (for the deterministic case) and @Reed1979 (under stochastic recruitment).
In short, the manager seeks to optimize the net present value (discounted cumulative catch) of a fishery, observing the stock size each year and setting an appropriate harvest quota in response.
In the classical approach, the best model of the fish population dynamics must first be estimated from data, potentially with posterior distributions over parameter estimates reflecting any uncertainty.
From this model, the optimal harvest policy -- that is, the a function which returns the optimal quota for each possible observed stock size -- can be determined either by analytic [@Reed1979] or numerical [@Marescot2013] methods, depending on the complexity of the model.
In a RL approach, the model is instead replaced by a computer simulation, usually referred to as an "environment."
The simplest simulation is merely an implementation of the estimated model itself.
However, unlike an analytic or SDP solution that is derived directly from the estimated model, the RL approach is in principle agnostic to the details of the simulation.

@@

## Ecological Tipping Points
